Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2022-09-07 paper/paper-infer.tex
discu

[KEEP RG FOR DISCU + CITE EXISTING METHODS]

[BETTER SET OF VARIANTS, BUT STILL NOT PERFECT E.G. FOR PROTEIN LEVELS]

[ALSO TALK ABOUT IMPUTATION]

[ALSO TALK ABOUT LDSC?]

[FUTURE WORK + LIMITATION: OVERESTIMATED H2 WITH SMALL LD + NEED TO WORK ON CORRECTION]

[ASSORTATIVE MATING FOR HEIGHT?]

[VERY POLYGENIC, BUT NOT INFINITESIMAL -> THEREFORE LDPRED2-INF NEVER RECOMMENDED (LDPRED2 + MISSPEC)]

[HM3\_PLUS AND MISSING HERITABILITY]

[MENTION YI'S PAPER]

[COMPARE FA ENRICHMENT TO YENGO2022]

@@

Here we introduce a few changes to LDpred2-auto, which makes it better at inferring these important parameters.
First, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
\begin{equation}\label{eq:new_model}
\beta_j = S_j \gamma_j \sim \left\{
\begin{array}{ll}
\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
0 & \mbox{otherwise.}
\end{array}
\right.
\end{equation}
Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite{zeng2021widespread}. 
In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before computing the maximum likelihood estimators, such that we add some proper sampling to these two parameters. 
This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration.
We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling by properly accounting for the reduced effective number of correlated variants.
As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler. 
We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.
Note that this $h^2$ estimate can be restricted to e.g.\ variants from a single LD block to get estimates of local heritability.

@@

%\clearpage
%\vspace*{3em}

@@

>>>>

LDpred2-auto was originally developed for building polygenic scores \cite[]{prive2020ldpred2}.
Here we have extended the LDpred2-auto model and shown that it can be used to reliably infer genetic architecture parameters such as the (local) SNP-heritability, polygenicity (and per-variant probabilities of being causal), and the selection-related parameter $\alpha$.
It can also infer the predictive performance $r^2$ of the resulting PGS (for the same genetic ancestry as the GWAS used for training).
We remind readers that it can also be used to infer the incertainty of individual polygenic scores \cite[]{ding2022large}.
Results across 248 phenotypes demonstrate that most of these phenotypes are very polygenic, yet do not have an infinitesimal architecture (i.e.\ not all variants are causal); this is consistent with LDpred2-inf generally providing much lower predictive performance than LDpred2-grid or LDpred2-auto \cite[]{prive2020ldpred2}.
We also obtain widespread signatures of selection with most $\alpha$ estimates between -1.0 and -0.4, consistent with previous findings \cite[]{zeng2021widespread}.
However, when looking at the heritability enrichment of several functional annotations for height, we obtain much smaller magnitudes than stratified LD Score regression \cite[]{finucane2018heritability}. For example, \cite{yengo2022saturated} report fold enrichments of more than 10x for e.g.\ coding and conserved variants, while we get less than 2x (Figure \ref{fig:enrichment}).

Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
However, we note that we are able to achieve an $r^2$ of 0.67 [TODO: UPDATE NUMBER] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.

Our proposed method have limitations. First, when power is low, estimates of $\alpha$ and $p$ become less reliable. However, estimates of $h^2$ and $r^2$ seem always reliable, except for maybe height for which they are probably overestimated.
We think this is due to assortative mating making the estimates too large in the presence of many spouses such as for the UK Biobank [TODO: CITATIONS? YENGO2022? OTHERS?]. 
Moreover, the $h^2$ from LDpred2-auto is also slightly overestimated when using a small LD reference panel or when the reference panel does not closely match with the ancestry of the GWAS summary statistics; future work could focus on correcting this issue. [TODO: OTHERS?]

Nevertheless, LDpred2-auto users can now get much from running a single method only. The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions. 
As future research directions, we are interested in using it for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, and multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}.

@@

Here we introduce a few changes to LDpred2-auto, which makes it better at inferring these important parameters.
First, we extend LDpred2-auto with a third parameter $\alpha$ that controls the relationship between minor allele frequencies (or equivalently, standard deviations) of genotypes and expected effect sizes; the model becomes
\begin{equation}\label{eq:new_model}
\beta_j = S_j \gamma_j \sim \left\{
\begin{array}{ll}
\mathcal N \big( 0,~\sigma_\beta^2 \cdot (S_j^2)^{(\alpha + 1)} \big) & \mbox{with probability $p$,} \\
0 & \mbox{otherwise.}
\end{array}
\right.
\end{equation}
Therefore, it was earlier assumed that $\alpha = -1$ and $\sigma_\beta^2 = h^2 / (M p)$ in equation \eqref{eq:prev_model}. 
This new model in equation \eqref{eq:new_model} is similar to the model assumed by SBayesS, where $\alpha$ is called $S$ \cite[]{zeng2021widespread}. 
In SBayesS, they estimate $\alpha$ and $\sigma_\beta^2$ by maximizing the likelihood of the normal distribution (over the causal variants from the Gibbs sampler).
In the new LDpred2-auto, we first sample causal variants with replacement (bootstrap) before computing the maximum likelihood estimators, such that we add some proper sampling to these two parameters. 
This maximum likelihood estimation is implemented using R package roptim \cite[]{pan2020roptim}, and we bound the estimate of $\alpha$ to be between -1.5 and 0.5 (the default, but can be modified), and the estimate of $\sigma_\beta^2$ to be between 0.5 and 2 times the one from the previous iteration.
We now sample $p$ from $\text{Beta}(1 + M_c / \bar{l^2}, 1 + (M - M_c) / \bar{l^2})$, where $\bar{l^2}$ is the average LD score, to add more sampling by properly accounting for the reduced effective number of correlated variants.
As for $h^2$, we still estimate it by $h^2 = \boldsymbol{\beta}^T \boldsymbol{R} \boldsymbol{\beta}$, where $\boldsymbol{R}$ is the correlation matrix between variants and $\boldsymbol{\beta}$ is a vector of causal effect sizes (after scaling) from one iteration of the Gibbs sampler. 
We constrain this estimate to be at least 0.001 to prevent the Gibbs sampler from being trapped in very small heritability estimates.
Note that this $h^2$ estimate can be restricted to e.g.\ variants from a single LD block to get estimates of local heritability.

@@

@@

\noindent[TODO: UPDATE LDPRED2 TUTORIAL TO INCLUDE INFERENCE]
